---
title: "Epicurve short presentation"
author: "Giovanni Marini"
date: "2024-04-10"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Background

This is a presentation for the epiCurve set of functions, which can be used to interpolate epidemiological curves assuming that the lying force of infection (FOI, i.e. the rate at which susceptible individuals acquire the infection) can be modelled through the density function of a normal distribution. There are two main assumptions beneath our analysis:

1. The infection is implicitly modelled by a Susceptible-Infected-Susceptible (SIS) model with an external FOI. Humans are therefore dead-end hosts and do not actively impact on the pathogen circulation.

2. The FOI has a seasonal pattern. This is the case of many vector-borne pathogens at temperate latitudes.

This modelling approach, introduced in (Marini et al. 2022), can therefore be used in the case of pathogens such as West Nile virus (WNV) or tick-borne encephalitis (TBE). It allows to retrospectively evaluate some key epidemiological parameters, namely when the FOI peaks, the length of the epidemiological season and the magnitude. 

# Modelling details
Let us assume we are considering some epidemiological data gathered between $n$ years $Y_1, ..., Y_n$ in some regions. We denote by $h_{y,i}(w)$ the number of recorded cases with region $i$ as place of infection with symptoms onset occurred during week $w$ of year $y$ ($w\in\{1, …, 52\}$, $y\in\{Y_1, ..., Y_n\}$), by $H_{y,i}$ the whole time series, i.e. $H_{y,i}=\bigcup_{w=1}^{52} h_{y,i}(w)$, and by $\sum_{y,i}$ the total number of cases with place of infection identified as i recorded during year y, i.e. $\sum_{y,i}=\sum_{w=1}^{52}h_{y,i}(w)$.

For the analysis, we assume $h_{y,i}(w)$ belongs to a Poisson distribution with average $\sum_{t\in T_W} N_i \cdot \lambda_{y,i}(t)$, where $\lambda_{y,i}(t)$ denotes the FOI in region $i$ and year $y$ at day $t$, $T_W$ represents the set of days in week $w$ and $N_i$ is the number of inhabitants of the region. The FOI is therefore 
$$\lambda_{y,i} (t)=c_{y,i}\cdot\frac{1}{\sigma_{y,i}\sqrt{2\pi}}e^{-\frac{1}{2}\left( \frac{t-\mu_{y,i}}{\sigma_{y,i}}\right)^2}.$$
Where $\mu_{y,i}$ and $\sigma_{y,i}$ represent respectively the average and standard deviation of the distribution and $c_{y,i}$ is a magnitude rescaling factor, representing the overall intensity of infections in that year and geographical area. These three parameters will be estimated by matching the generated epidemiological curve to the observed data through a maximum likelihood approach. 

For further additional details please see (Marini et al. 2022).

# Loading data
First of all we load the `tidyverse` library, the functions needed to carry out the analysis and our files. More specifically, we need two files, one CSV with the case-based data and another one with the number of inhabitants for each considered region.
```{r beginning, message = FALSE}
library(tidyverse)
source("epiCurve_functions.R")
cases_df=read.csv("Data/disease_cases.csv")
population=read.csv("data/population.csv")
```
Let's see how the two CSV files should be:
```{r files_inspection}
head(cases_df)
head(population)
```
So each line of the epidemiological file represents a single case with their region of infection (in this example denoted by a letter, but it could be a municipality, a province, a country...) and the date of the onset of the symptoms.

The population file consists of the number of inhabitants for each region.

We can plot our data to see the total number of cases during the period of interest.

```{r preliminary_plots, include=T, out.width="100%", fig.align = 'center'}
cases_df$ONSET_DATE=as.Date(cases_df$ONSET_DATE)

my_theme=theme(panel.background = element_blank(),
               axis.title=element_text(size=20), axis.text=element_text(size=16))

hist_week=ggplot(data=cases_df)+
  geom_bar(aes(x=ONSET_DATE))+
  labs(x="Date of symptoms onset", y="Count")+
  my_theme

hist_week
```

So cases are recorded seasonally, which is one of the major requirements for our modelling approach.

# Modelling
We are now going to fit each epidemiological curve $H_{y,i}$ , i.e. each set of cases recorded during each year and each region. This is done through the `epi_fit` function, after preparing the appropriate dataset through the `create_dataset` function. In this latter function we specify the name of the raw dataset and the minimum number of total cases for each epidemiological curve to be considered (i.e., we will model only those curves with $\sum_{y,i}\geq$`minimum_cases`). For our example we set `minimum_cases=5`.

The variable `run_fit` specified whether we want to run the fit. If we already did this part we can set it to `FALSE` as we already have the output saved in the specified CSV file.

```{r model_fit, include=T}
output_file_name="normal_FOI_fit.csv"
minimum_cases=5
matrix_weekly_cases=create_dataset(minimum_cases, cases_df)
run_fit=F
if(run_fit)
  epi_fit(matrix_weekly_cases,population,output_file_name)
```

We can now load the output file and see what it contains:
```{r output_inspection}
output_model=read.csv("normal_FOI_fit.csv",header=T)
head(output_model)
```
So each line represents a particular epidemiological curve $H_{y,i}$ as specified by the region and year (first two columns). The third column indicates the number of inhabitants of the region, the fourth and the fifth provide the observed ($\sum_{y,i}$) and modelled ($\bar{\sum_{y,i}}$) total number of cases respectively. Finally in the last three columns the best likelihood estimated for the three FOI parameters are given: $\mu$, $\sigma$ and $c$.

First of all we can check how well we fitted the curves, for instance through a scatterplot and a correlation test:
```{r fit_quality, fig.align = 'center'}
cor.test(output_model$OBSERVED_TOTAL,output_model$MODEL_TOTAL)
plot(output_model$OBSERVED_TOTAL,output_model$MODEL_TOTAL,
     xlab="Observed total cases",ylab="Modelled total cases",
     xlim=c(minimum_cases,max(c(output_model$OBSERVED_TOTAL,output_model$MODEL_TOTAL))),
     ylim=c(minimum_cases,max(c(output_model$OBSERVED_TOTAL,output_model$MODEL_TOTAL))))
abline(a=0,b=1)
```

So there is quite a good agreement between the two quantities. We can also evaluate the average squared error as `mean((output_model$OBSERVED_TOTAL-output_model$MODEL_TOTAL)^2)` which in our case equals to `r round(mean((output_model$OBSERVED_TOTAL-output_model$MODEL_TOTAL)^2),1)`.

We can also visually compare an observed curve with its fit. For instance, let's see region a during 2016.

```{r curves_comparison, warning=FALSE, fig.align = 'center',out.width="100%"}
selected_year=2016
selected_region="a"
selected_row=which(matrix_weekly_cases$REGION==selected_region & matrix_weekly_cases$YEAR==selected_year)

weekly_observed_cases=as.numeric(matrix_weekly_cases[selected_row,-c(1,2)])
N=as.numeric(output_model$POPULATION[selected_row])
weekly_model_cases=c()
for(week_start_index in seq(1,364,7)){
  c_FOI=output_model$C_NORM[selected_row]
  mu_FOI=output_model$MU[selected_row]
  sd_FOI=output_model$SD[selected_row]
  FOI_model=c_FOI*dnorm(1:364, mean = mu_FOI, sd = sd_FOI)
  
  tmp=0
  for(day in week_start_index:(week_start_index+6))
    tmp=tmp+N*FOI_model[day]
  weekly_model_cases=c(weekly_model_cases,tmp)
}

ymax=max(weekly_observed_cases,weekly_model_cases)

df_to_plot=data.frame(WEEK=1:length(weekly_observed_cases),
                      CASES=weekly_observed_cases,TYPE="OBS")
df_to_plot=rbind(df_to_plot,data.frame(WEEK=1:length(weekly_observed_cases),
                                       CASES=weekly_model_cases,TYPE="MODEL"))
ggplot(df_to_plot,aes(x=WEEK,y=CASES))+
  geom_line(aes(x=WEEK, y=CASES,col=TYPE),size=1)+
  geom_point(aes(x=WEEK, y=CASES,col=TYPE,shape=TYPE),size=3)+
  labs(x = "Week", y=expression("h"["y,i"]))+
  my_theme+
  theme(axis.line=element_line(linewidth=1))+
  scale_color_manual(values=c("orchid2","mediumpurple2"))
```

# References
Marini, G., Pugliese, A., Wint, W., Alexander, N.S., Rizzoli, A., Rosà, R., 2022. Modelling the West Nile virus force of infection in the European human population. One Health 15, 100462. https://doi.org/10.1016/j.onehlt.2022.100462
